Continous time Markov chain models; \\ extension, simulation and visualisation

### 2014-09-30 chapters/chapter01.tex
change the figure per Philippes suggestion

% external software
BEAST comes with a number of additional external software packages that can either assist in generating the input for BEAST, analyzing the results or for some other more specific tasks.
\textbf{BEAUti} (Bayesian Evolutionary Analysis Utility) is a  GUI (graphical user-interface) application for generating XML input files for BEAST. 
% It alleviates some of the most cumbersome tasks that would otherwise have to be done manually, thus saving time and simplifying 
It allows for manipulation of models, priors, operators, and MCMC chain settings, without resorting to a manual editing of the XML syntax.
\textbf{LogCombiner} combines log files and tree log files from multiple runs, that can be than importer into \textbf{Tracer}, a graphical tool for the analysis and visualisation of the output of BEAST as well as other common MCMC packages.
It allows for a quick inspection of mixing, convergence, and distribution of the parameters of interest.
\textbf{TreeAnnotator} summarizes the tree log files as MCC (maximum clade credibility) trees and allows for annotating tree branches with external information, such as posterior probabilities, a posteriori most probable ancestral traits and other.
These information can then be visualised using \textbf{FigTree} or if nodes of the trees are annotated with spatial information they can be displayes in geographical coordinates using \textbf{SPREAD}.
Finally \textbf{PIBUSS} is a software for sequence simulation, capable of generating XML files understood by BEAST for combined simulation-inference analysis as well as sequences in formats taht can be imported into any other phylogenetic software.

@@

%GB: perhaps this subsection can be structured more clearly by using the \paragraph command for each of the 3 subtopics?

% models
Phylogenetic analyses employing stochastic processes make strong restrictive assumptions regarding the underlying substitution process.
These assumptions are mainly considered to reduce the number of free parameters of the models and ease the already mentioned computational and mathematical tractability.
Unfortunately many of these assumptions hamper the application of the methods to the complex, real-world data-sets.
Relaxing some of these assumptions and development of more biologically plausible models, able to grasp the complex nature of phylodynamic dispersal processes is one the aims of the presented thesis \citep{Bielejec2014a}.
Another goal is to make the general applicability suitable for epidemiological problems as well as beyond them, with potential to turn the findings into useful predictions of the emergence of infectious diseases, e.g. by identifying the sources and sinks of the viral spread.
% geographical dispersal

% simulation
With the development of novel phylogenetic inference methods comes the need to synthesize evolutionary data, in order to compare estimator performance and characterize strengths and weaknesses of different approaches (e.g. \cite{Arenas2012}, \cite{Hoban2011}).
Whereas the true underlying evolutionary relationships between biological sequences are generally unknown, Monte Carlo simulations allow generating test scenarios while controlling for the evolutionary history as well as the tempo and mode of evolution. 
Part of the research presented in this thesis is devoted towards development of a flexible software for sequence simulation, that integrates both the tree-generative processes, as well as models responsible for shaping the molecular histories and geographical dispersal \citep{Bielejec2014a}.

% visualisation
Phylogenetic inference inevitably results in complex, multi-dimensional output data.
This dimensionality is further incremented by adding the geography to the usual evolutionary time component.
Finally, Bayesian analyses result in posterior distributions of trees, with each tree branch annotated with estimated node locations and other traits of interest.
These aspects represent a major challenge for visualizing the phylogenetic inference.
Yet clear and intuitive visualizations are an important aspect of every analysis \citep{Hadley2010}.

@@

\begin{figure}[H]
\centering
\includegraphics[scale=0.3]{figures/spread}
\caption{
{ \footnotesize 
{\bf Phylogeographic visualization of viral spread in time and space.} 
A. Conceptual representation of time-slicing a phylogeny and projecting the part of the tree up to time $t_{1}$ and $t_{2}$ respectively. B. An example projection of influenza H5N1 over time in Google Earth for two different segment phylogenies. % (hemagglutinin and neuraminidase in green and magenta respectively).
} % END: footnotesize
}
\label{fig:spread}
\end{figure}

Presenting readily interpretable visual summaries of the inferences, that can in turn be communicated to the researchers across different fields is one of the more important aspects of this thesis.
This goal has been materialized in the development of a user-friendly application, that produces compelling and informative statistical visualizations of viral spread in time and space \citep{Bielejec2011}.
An example of such a visualisation can be found in Figure~\ref{fig:spread}.
%PL: never say 'there is a figure for you' without bothering to discuss it. At this point the reader has no idea what to make of time slicing, etc... This is an explanation of a SPREAD approach and does not belong here in the intro. Perhaps if you pick it up in the discussion, you could say that the paper format did not allow you to go into the visualization strategy all that much and you expand it on it a bit before discussing how you see the future for such things

@@

%PL: I refer to our conversation about this fig.
% TODO change that plot, show multiple subst per branch (and how it alters time estimates)
\begin{figure}[H]
\centering
\begingroup
\everymath{\displaystyle}
{\Large
\begin{displaymath} % 
    \xymatrix@R+1pc{ 
\color{gray2}A \ar[r] & \color{gray2}T \ar[r] & A \\
\color{gray2}A \ar[r] & \color{gray2}T \ar[r] & G \\
\color{gray2}A  \ar[rr]_{ \color{gray2}\begin{array}{c}
A\rightarrow G\\
G\rightarrow C
\end{array}} &&  C
    } % 
\end{displaymath}
}% END: Large
\endgroup
\caption{
{ \footnotesize 
{\bf Evolution at a single site.} Over time characters at the site change, leading to the observed data (black). Some of the mutations may be silent. Each site evolves independently.
}% END: footnotesize
}
\label{fig:alignment}
\end{figure}

>>>>

% external software
BEAST comes with a number of additional external software packages that can either assist in generating the input for BEAST, analyzing the results or for some other more specific tasks.
\textbf{BEAUti} (Bayesian Evolutionary Analysis Utility) is a  GUI (graphical user-interface) application for generating XML input files for BEAST. 
% It alleviates some of the most cumbersome tasks that would otherwise have to be done manually, thus saving time and simplifying 
It allows for manipulation of models, priors, operators, and MCMC chain settings, without resorting to a manual editing of the XML syntax.
\textbf{LogCombiner} combines log files and tree log files from multiple runs, that can be than importer into \textbf{Tracer}, a graphical tool for the analysis and visualisation of the output of BEAST as well as other common MCMC packages.
It allows for a quick inspection of mixing, convergence, and distribution of the parameters of interest.
\textbf{TreeAnnotator} summarizes the tree log files as MCC (maximum clade credibility) trees and allows for annotating tree branches with external information, such as posterior probabilities, a posteriori most probable ancestral traits and other.
These information can then be visualised using \textbf{FigTree} or if nodes of the trees are annotated with spatial information they can be displayes in geographical coordinates using \textbf{SPREAD}.
Finally \textbf{$\pi$BUSS} is a software for sequence simulation, capable of generating XML files understood by BEAST for combined simulation-inference analysis as well as sequences in formats taht can be imported into any other phylogenetic software.

@@

% models
\paragraph{}
Phylogenetic analyses employing stochastic processes make strong restrictive assumptions regarding the underlying substitution process.
These assumptions are mainly considered to reduce the number of free parameters of the models and ease the already mentioned computational and mathematical tractability.
Unfortunately many of these assumptions hamper the application of the methods to the complex, real-world data-sets.
Relaxing some of these assumptions and development of more biologically plausible models, able to grasp the complex nature of phylodynamic dispersal processes is one the aims of the presented thesis \citep{Bielejec2014a}.
Another goal is to make the general applicability suitable for epidemiological problems as well as beyond them, with potential to turn the findings into useful predictions of the emergence of infectious diseases, e.g. by identifying the sources and sinks of the viral spread.
% geographical dispersal

% simulation
\paragraph{}
With the development of novel phylogenetic inference methods comes the need to synthesize evolutionary data, in order to compare estimator performance and characterize strengths and weaknesses of different approaches (e.g. \cite{Arenas2012}, \cite{Hoban2011}).
Whereas the true underlying evolutionary relationships between biological sequences are generally unknown, Monte Carlo simulations allow generating test scenarios while controlling for the evolutionary history as well as the tempo and mode of evolution. 
Part of the research presented in this thesis is devoted towards development of a flexible software for sequence simulation, that integrates both the tree-generative processes, as well as models responsible for shaping the molecular histories and geographical dispersal \citep{Bielejec2014a}.

% visualisation
\paragraph{}
Phylogenetic inference inevitably results in complex, multi-dimensional output data.
This dimensionality is further incremented by adding the geography to the usual evolutionary time component.
Finally, Bayesian analyses result in posterior distributions of trees, with each tree branch annotated with estimated node locations and other traits of interest.
These aspects represent a major challenge for visualizing the phylogenetic inference.
Yet clear and intuitive visualizations are an important aspect of every analysis \citep{Hadley2010}.

@@

%TODO: move to discussion

% \begin{figure}[H]
% \centering
% \includegraphics[scale=0.3]{figures/spread}
% \caption{
% { \footnotesize 
% {\bf Phylogeographic visualization of viral spread in time and space.} 
% A. Conceptual representation of time-slicing a phylogeny and projecting the part of the tree up to time $t_{1}$ and $t_{2}$ respectively. B. An example projection of influenza H5N1 over time in Google Earth for two different segment phylogenies. % (hemagglutinin and neuraminidase in green and magenta respectively).
% } % END: footnotesize
% }
% \label{fig:spread}
% \end{figure}

Presenting readily interpretable visual summaries of the inferences, that can in turn be communicated to the researchers across different fields is one of the more important aspects of this thesis.
This goal has been materialized in the development of a user-friendly application, that produces compelling and informative statistical visualizations of viral spread in time and space \citep{Bielejec2011}.
% An example of such a visualisation can be found in Figure~\ref{fig:spread}.
%PL: never say 'there is a figure for you' without bothering to discuss it. At this point the reader has no idea what to make of time slicing, etc... This is an explanation of a SPREAD approach and does not belong here in the intro. Perhaps if you pick it up in the discussion, you could say that the paper format did not allow you to go into the visualization strategy all that much and you expand it on it a bit before discussing how you see the future for such things

@@

\begin{figure}[H]
\centering
\begingroup
\everymath{\displaystyle}
{\Large
\begin{displaymath} % 
\xymatrix{
& \color{gray2}A \ar[dl] \ar[drdr]^{ \color{gray2}\begin{array}{c}
A\rightarrow G\\
G\rightarrow C
\end{array}} & & \\
T & & & \\
& & & C \\
}
%     \xymatrix@R+1pc{ 
% \color{gray2}A \ar[r] & \color{gray2}T \ar[r] & A \\
% \color{gray2}A \ar[r] & \color{gray2}T \ar[r] & G \\
% \color{gray2}A  \ar[rr]_{ \color{gray2}\begin{array}{c}
% A\rightarrow G\\
% G\rightarrow C
% \end{array}} &&  C
%     } % 
\end{displaymath}
}% END: Large
\endgroup
\caption{
{ \footnotesize 
{\bf Evolution at a single site.} Over time characters at the site change, leading to the observed data (black). Some of the mutations may be silent. Each site evolves independently.
}% END: footnotesize
}
\label{fig:alignment}
\end{figure}
